---
title: "Single-cell RNA sequencing ~ Session 1<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "http://rockefelleruniversity.github.io/RU_course_template/"
output: 
  xaringan::moon_reader:
    css: ["default", "metropolisCustom.css", "metropolis-fontsCustom.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
  html_document:
    toc: true # table of content true
    toc_float: yes
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
params:
  isSlides: "no"
---

```{r,include=FALSE}
suppressPackageStartupMessages(require(knitr))
knitr::opts_chunk$set(echo = TRUE, tidy = T)
```

---
## Overview
In this course we are going to introduce basic analysis for single-cell RNAseq, with a specific focus on the 10X system. The course is divided into three sessions. In the first session, we will introduce how to interpret the Cell  Ranger QC report and how to do analysis with the LOUPE Browser. We will also demonstrate the customized analysis we can support right know. In the Second session, we will demonstrate how to process scRNAseq data and make QC reports with Seurat. In the third session, we will discuss how to to do more advanced analysis and QC with Bioconductor packages. 

---
```{r, results='asis',include=TRUE,echo=FALSE}
if(params$isSlides == "yes"){
  cat("class: inverse, center, middle

# Cell Ranger results

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=720px></html> 

---
"    
  )
}else{
  cat("# Cell Ranger results


---
"    
  )
  
}

```

## Cell Ranger
[Cell Ranger](https://support.10xgenomics.com/single-cell-gene-expression/software/pipelines/latest/what-is-cell-ranger) is a widely used software for single-cell sequencing data analysis. It supports various kinds of analysis for 10X single-cell sequencing data.

- Alignment and counting
- Analysis of VDJ changes
- Datasets aggregation
- Convert BCL files (raw files) into FASTQ files
- Generation of customized GTF and Genome files


---
## Basic reports of Cell Ranger
- [QC Report]("data/WT.html")
- cloupe file: can be processed by [LOUPE Browser](https://support.10xgenomics.com/single-cell-gene-expression/software/visualization/latest/what-is-loupe-cell-browser)


---
```{r, results='asis',include=TRUE,echo=FALSE}
if(params$isSlides == "yes"){
  cat("class: inverse, center, middle

# Customized Analysis

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=720px></html> 

---
"    
  )
}else{
  cat("# Customized Analysis


---
"    
  )
  
}

```

## Demonstration
Here we use an article published by Fuchs Lab as an example to demonstrate the customized analysis we have done here at the BRC. Please access this [article](https://elifesciences.org/articles/56980). 
- General QC plots
- Data integration and normalization
- Clustering
- Annotate cell types
- Pseudotime analysis
- Differential gene expression and single-cell pathway analysis

---
## The next two sessions
We will also use the data of the eLife paper in next two sessions. You can download the data from GEO [link](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE147415) or from BOX [link](https://rockefeller.box.com/s/hdy0vex38dlph5pglafrkw68b7uemndr).
